Measurement Clusterization in D-optimal Designs for Bayesian Linear Inverse Problems over Hilbert Spaces

### 2023-05-21 arxiv/intro.tex
working on arxiv
Ready for arxiv

Surprisingly, A- and D-optimal designs have been observed to yield
remarkably similar measurements in certain cases \cite{fedorov1996,
  hooker2009, fedorov1997, Ucinski05, neitzel2019sparse}. To
illustrate this point, we consider a toy inverse problem: inferring
the initial condition of the 1D heat equation (details in the
supplementary material). In Figure \ref{fig:clusterization
  illustration} D-optimal measurement locations are shown for
different numbers of measurements. Notably, for six measurements, a
D-optimal design yields two sets of measurements that are almost
indistinguishable from one another. We refer to this intriguing
phenomenon as \emph{measurement clusterization} \cite{Ucinski05}, and
we refer to a design that exhibits measurement clusterization as a
\emph{clustered design}.

%% first: blue, second: red, third: green, fourth: black,
%% fifth:magenta
\begin{figure}[H]
    \centering
    \includegraphics[height=0.5\textwidth]{example.eps}
    \caption{Measurement clusterization for optimal designs when
      inverting for the initial condition of the 1D heat equation (see
      supplementary material for details). Measurement locations were
      chosen according to the Bayesian D-optimality criterion of
      Theorem \ref{thm:d optimality}. Measurement locations are
      plotted over the computational domain $\Omega = [0, 1]$
      (x-axis), for varying numbers of measurements (y-axis). The
      colored numbers are measurement indices, plotted for visual
      clarity. Measurement clusterization already occurs for three
      measurements: the second measurement (red) is overlaid with the
      third (green). For five measurements, first (blue) and second
      (red) measurements are clustered, as well as the fourth (black)
      and the fifth (magenta).}
  \label{fig:clusterization illustration}
\end{figure}

Researchers widely agree that measurement clusterization is
undesirable \cite{fedorov1996, hooker2009, fedorov1997, Ucinski05,
  neitzel2019sparse}, prompting the exploration of various remedies to
address this issue. One approach involves merging close measurements
\cite{fedorov1997}; however, this strategy merely overlooks the
phenomenon of measurement clusterization. An alternative solution lies
in \emph{clusterization-free design}s, where measurement locations are
deliberately chosen to be distant from one another. This can be
achieved by imposing distance constraints between measurements or by
introducing correlated errors that account for both observation error
and model misspecification \cite{Ucinski05}. For instance, in the
context of time-series analysis for pharmacokinetic experiments,
measurement clusterization can be mitigated by incorporating the
modeling of auto-correlation time within the noise terms
\cite{hooker2009}.

>>>>

Surprisingly, A- and D-optimal designs have been observed to yield
remarkably similar measurements in certain cases \cite{fedorov1996,
  nyberg2012, fedorov1997, Ucinski05, neitzel2019sparse}. To
illustrate this point, we consider a toy inverse problem: inferring
the initial condition of the 1D heat equation (details in the
supplementary material). In Figure \ref{fig:clusterization
  illustration} D-optimal measurement locations are shown for
different numbers of measurements. Notably, for six measurements, a
D-optimal design yields two sets of measurements that are almost
indistinguishable from one another. We refer to this intriguing
phenomenon as \emph{measurement clusterization} \cite{Ucinski05}, and
we refer to a design that exhibits measurement clusterization as a
\emph{clustered design}.

%% first: blue, second: red, third: green, fourth: black,
%% fifth:magenta
\begin{figure}
    \centering
    \includegraphics[height=0.5\textwidth]{example.pdf}
    \caption{Measurement clusterization for optimal designs when
      inverting for the initial condition of the 1D heat equation (see
      supplementary material for details). Measurement locations were
      chosen according to the Bayesian D-optimality criterion of
      Theorem \ref{thm:d optimality}. Measurement locations are
      plotted over the computational domain $\Omega = [0, 1]$
      (x-axis), for varying numbers of measurements (y-axis). The
      colored numbers are measurement indices, plotted for visual
      clarity. Measurement clusterization already occurs for three
      measurements: the second measurement (red) is overlaid with the
      third (green). For five measurements, first (blue) and second
      (red) measurements are clustered, as well as the fourth (black)
      and the fifth (magenta).}
  \label{fig:clusterization illustration}
\end{figure}

Researchers widely agree that measurement clusterization is
undesirable \cite{fedorov1996, nyberg2012, fedorov1997, Ucinski05,
  neitzel2019sparse}, prompting the exploration of various remedies to
address this issue. One approach involves merging close measurements
\cite{fedorov1997}; however, this strategy merely overlooks the
phenomenon of measurement clusterization. An alternative solution lies
in \emph{clusterization-free design}s, where measurement locations are
deliberately chosen to be distant from one another. This can be
achieved by imposing distance constraints between measurements or by
introducing correlated errors that account for both observation error
and model misspecification \cite{Ucinski05}. For instance, in the
context of time-series analysis for pharmacokinetic experiments,
measurement clusterization can be mitigated by incorporating the
modeling of auto-correlation time within the noise terms
\cite{nyberg2012}.

### 2023-05-21 arxiv/ms.tex
working on arxiv
Ready for arxiv

\documentclass{siamonline220329}
%\documentclass{amsart}
%\documentclass{article}
\input{definitions}

@@

\title{Measurement Clusterization in D-optimal Designs for Bayesian Linear
  Inverse Problems over Hilbert Spaces}

\author{Yair Daon\thanks{Porter School of the Environment and Earth
    Sciences, Tel Aviv University, Tel Aviv, Israel
    (\email{yair.daon@gmail.com}).}}
%% \author{Yair Daon\thanks{Azrieli Faculty of Medicine, Bar Ilan
%%     University, Safed, Israel.  (\email{yair.daon@gmail.com}).}}

%% \address{Porter School of the Environment and Earth
%%   Sciences, Tel Aviv University\\ Tel Aviv, Israel}
%% %% \curraddr{Porter School of Environment Studies, Tel Aviv University\\ Tel Aviv, Israel}
%% %% \curraddr{}
%% \email{yair.daon@gmail.com}
% The 2010 edition of the Mathematics Subject Classification is the
% current definitive version.
%% \subjclass{Primary: 
%%   62F15, % Statistics - bayesian inference
%%   35R30, % PDE - Inverse problems
%%   Secondary:
%%   28C20,
%%   % Measure and integration - set functions and measures and integrals
%%          % in infinite-dimensional spaces (Wiener measure, Gaussian
%%   % measure, etc.
%% }

\date{\today}
\begin{document}
\maketitle
\begin{abstract}
  Inversion for parameters of physical processes typically requires
  taking expensive measurements; finding an optimal set of
  measurements is known as the optimal design problem. Surprisingly,
  measurement locations in optimal designs are sometimes extremely
  clustered. Since measurement clusterization is nonintuitive,
  researchers often avoid it by modifying the optimal design
  problem. We consider a certain flavor of the optimal design problem,
  based on the Bayesian D-optimality criterion, and suggest an
  analytically tractable model for Bayesian D-optimal designs in
  linear inverse problems over Hilbert spaces. We demonstrate that
  measurement clusterization is a generic property of D-optimal
  designs, and prove that correlated noise between measurements
  mitigates clusterization. We also give a full characterization of
  D-optimal designs, which shows that D-optimal designs uniformly
  reduce uncertainty in a select subset of the prior covariance
  eigenvectors. Finally, we show how measurement clusterization is a
  simple consequence of the characterization mentioned above and the
  pigeonhole principle.
\end{abstract}

\maketitle

@@

\bibliographystyle{siamplain}
%% %%\bibliographystyle{amsplain}
\bibliography{/home/yair/projects/bibtex.bib}

>>>>

%\documentclass{siamonline220329}
\documentclass{amsart}
\input{definitions}

@@

\title[Measurement Clusterization in D-optimal Designs]{Measurement
  Clusterization in D-optimal Designs for Bayesian Linear Inverse
  Problems over Hilbert Spaces}

\author{Yair Daon}

%% \thanks{Porter School of the Environment and Earth
%%     Sciences, Tel Aviv University, Tel Aviv, Israel
%%     (\email{yair.daon@gmail.com}).}
%% \author{Yair Daon\thanks{Azrieli Faculty of Medicine, Bar Ilan
%%     University, Safed, Israel.  (\email{yair.daon@gmail.com}).}}

\address{Porter School of the Environment and Earth Sciences, Tel Aviv
  University\\ Tel Aviv, Israel}
\address{Sackler Faculty of Medicine,
  Sciences, Tel Aviv University\\ Tel Aviv, Israel}

%% %% \curraddr{Porter School of Environment Studies, Tel Aviv University\\ Tel Aviv, Israel}
%% %% \curraddr{}
\email{yair.daon@gmail.com}
%% The 2010 edition of the Mathematics Subject Classification is the
%% current definitive version.

\subjclass{Primary: 
  62F15, % Statistics - bayesian inference
  35R30, % PDE - Inverse problems
  Secondary:
  28C20.
  % Measure and integration - set functions and measures and integrals
         % in infinite-dimensional spaces (Wiener measure, Gaussian
  % measure, etc.
}

\date{\today}
\begin{document}

\begin{abstract}
  Inversion for parameters of physical processes typically requires
  taking expensive measurements; finding an optimal set of
  measurements is known as the optimal design problem. Surprisingly,
  measurement locations in optimal designs are sometimes extremely
  clustered. Since measurement clusterization is nonintuitive,
  researchers often avoid it by modifying the optimal design
  problem. We consider a certain flavor of the optimal design problem,
  based on the Bayesian D-optimality criterion, and suggest an
  analytically tractable model for Bayesian D-optimal designs in
  linear inverse problems over Hilbert spaces. We demonstrate that
  measurement clusterization is a generic property of D-optimal
  designs, and prove that correlated noise between measurements
  mitigates clusterization. We also give a full characterization of
  D-optimal designs, which shows that D-optimal designs uniformly
  reduce uncertainty in a select subset of the prior covariance
  eigenvectors. Finally, we show how measurement clusterization is a
  simple consequence of the characterization mentioned above and the
  pigeonhole principle.
\end{abstract}
\maketitle

@@

%\bibliographystyle{siamplain}
\bibliographystyle{amsplain}
\bibliography{/home/yair/projects/bibtex.bib}

### 2023-05-21 arxiv/prelims.tex
Ready for arxiv

It is useful to record that the posterior measure $\post$ is Gaussian
in this setting and its covariance operator is \cite{Stuart10}:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}
Note that the poserior covariance $\postcov$ does not depend on data
$\data$.

@@

The study of D-optimal designs for Bayesian linear inverse problems in
infinite dimensions was pioneered by \cite{AlexanderianGloorGhattas14,
  alexanderian2018efficient}. The main result we will make use of is
summarized (in our notation) in Theorem \ref{thm:d optimality}:

@@

Note that in \cite{AlexanderianGloorGhattas14,
  alexanderian2018efficient}, results are stated for $\Sigma=I$
(implied by $\modcov = 0,\sigma^2=1$), but the authors also mention
these results hold for more general covariance matrices
\cite[p. 681]{AlexanderianGloorGhattas14}.

%% It is important to note that since $\obs$ is finite-rank,
%% $\prcov^{1/2} \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}$ is
%% trace-class.
\begin{definition}\label{def:d optimality}
  $\obs^{\star}$ is said to be \emph{D-optimal} if $\obs^{\star} =
  \argmax_{\obs} \tar(\obs)$, where entries of $\obs \in (\hilo^*)^m$
  are constrained to some allowed set of measurements in $\hilo^*$.
\end{definition}

For a Bayesian linear model in finite dimensions, with Gaussian prior
and Gaussian noise, a D-optimal design minimizes the determinant of
the posterior covariance matrix, and this turns out to be a
regularized version of the frequentist criterion
\cite{Chaloner1995}. Theorem \ref{thm:d optimality} and Definition
\ref{def:d optimality} carry a similar intuition:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \frac12 \log \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \text{ by \eqref{eq:objective}}\\
    &= \frac12 \log \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \frac12 \log \det \prcov \postcov^{-1} \text{ by \eqref{eq:postcov}}.
    %%  &= \frac12 \log \det \prcov -\frac12 \log \det \postcov.
  \end{split}
\end{align*}
We think of $\prcov$ as constant, so a D-optimal design minimizes the
posterior covariance determinant, analogously to the
finite-dimensional case.

>>>>

Finally, it is useful to record that the posterior measure $\post$ is
Gaussian in this setting and its covariance operator is does not
depend on data $\data$ \cite{Stuart10}:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}

@@

The study of D-optimal designs for Bayesian linear inverse problems in
infinite dimensions was pioneered by \cite{AlexanderianGloorGhattas14,
  alexanderian2018efficient}. The main result we will make use of is
summarized in our notation below:

@@

Note that in \cite{AlexanderianGloorGhattas14,
  alexanderian2018efficient}, results are stated for $\Sigma=I$
(implied by $\modcov = 0,\sigma^2=1$), but the these results also hold
for more general covariance matrices
\cite[p. 681]{AlexanderianGloorGhattas14}.

%% It is important to note that since $\obs$ is finite-rank,
%% $\prcov^{1/2} \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}$ is
%% trace-class.
\begin{definition}\label{def:d optimality}
  We say $\obs^{\star}$ is \emph{D-optimal} if $\obs^{\star} =
  \argmax_{\obs} \tar(\obs)$, where entries of $\obs \in (\hilo^*)^m$
  are constrained to some allowed set of measurements in $\hilo^*$.
\end{definition}

For a Bayesian linear model in finite dimensions, with Gaussian prior
and Gaussian noise, a D-optimal design minimizes the determinant of
the posterior covariance matrix, and this turns out to be a
regularized version of the frequentist criterion
\cite{Chaloner1995}. Theorem \ref{thm:d optimality} and Definition
\ref{def:d optimality} carry a similar intuition:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \frac12 \log \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \text{ by \eqref{eq:objective}}\\
    &= \frac12 \log \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \frac12 \log \det \prcov \postcov^{-1} \text{ by \eqref{eq:postcov}}.
    %%  &= \frac12 \log \det \prcov -\frac12 \log \det \postcov.
  \end{split}
\end{align*}
We think of $\prcov$ as constant, so a D-optimal design minimizes a
quantity analogous to the posterior covariance determinant, similarly
to the finite-dimensional case.

### 2023-05-21 arxiv/vanishing.tex
Ready for arxiv

\section{Characterization of D-Optimal Designs when $\modcov = 0$}\label{section:vanishing}
Our goal in this section is to prove Theorem \ref{thm:char} which
characterizes D-optimal designs when $\modcov = 0$. The necessary
first-order condition for D-optimality of Theorem
\ref{thm:constrained} for $\modcov = 0$ become:

@@

The final ingredient we require for the proof of Theorem
\ref{thm:char} the following lemma:

@@

\begin{figure}
  \begin{tikzpicture}[scale=0.85]
    \begin{axis}[
        ybar stacked,
        ymin=0,
        ymax=4,
        xtick=data,
        legend style={cells={anchor=east}, legend pos=north west, legend columns=-1},
        reverse legend=false, % set to false to get correct display, but I'd like to have this true
        xticklabels from table={\optimalvsnot}{Label},
        xticklabel style={text width=2cm,align=center},
        legend plot pos=right,
        ylabel=precision --- prior and posterior,
        xlabel=eigenvector $i$,
      ]

      \addplot [fill=green!80]  table [y=prior,   meta=Label, x expr=\coordindex] {\optimalvsnot};
      \addplot [fill=blue!60]   table [y=optimal, meta=Label, x expr=\coordindex] {\optimalvsnot};

      \addlegendentry{$\sigma^2\lambda_i^{-1}$}
      \addlegendentry{optimal $\eta_i$s}

    \end{axis}
  \end{tikzpicture}
  \qquad
  \begin{tikzpicture}[scale=0.85]
    \begin{axis}[
        ybar stacked,
        ymin=0,
        ymax=4,
        xtick=data,
        legend style={cells={anchor=east}, legend pos=north west, legend columns=-1},
        reverse legend=false, % set to false to get correct display, but I'd like to have this true
        xticklabels from table={\optimalvsnot}{Label},
        xticklabel style={text width=2cm,align=center},
        legend plot pos=right,
        ylabel=precision --- prior and posterior,
        xlabel=eigenvector $i$,
      ]

      \addplot [fill=green!80]  table [y=prior,       meta=Label, x expr=\coordindex] {\optimalvsnot};
      \addplot [fill=blue!60]   table [y=sub-optimal, meta=Label, x expr=\coordindex] {\optimalvsnot};

      \addlegendentry{$\sigma^2\lambda_i^{-1}$}
      \addlegendentry{sub-optimal $\eta_i$s}

    \end{axis}
  \end{tikzpicture}
  \caption{A comparison of the eigenvalues of the pushforward
    posterior precision $(\fwd\prcov\fwd^*)^{-1} +
    \sigma^{-2}\obs^*\obs$ for a D-optimal design (left) and a
    sub-optimal design (right). Both designs are allowed $m=3$
    measurements. We assume $\sigma^2=1$ and thus, the blue area has
    accumulated height of $\sigma^{-2}m = 3$ in both panels. The
    D-optimal design (left) increases precision where it is
    lowest. The sub-optimal design (right) does not.}
  \label{fig:optimal vs not}
\end{figure}

>>>>

\section{D-Optimal Designs Without Model Error}\label{section:vanishing}
Our goal in this section is to prove Theorem \ref{thm:char} which
characterizes D-optimal designs when $\modcov = 0$. The necessary
first-order condition for D-optimality of Theorem
\ref{thm:constrained} for $\modcov = 0$ become:

@@

The final ingredient we require for the proof of Theorem
\ref{thm:char} is:

@@

\begin{figure}
  \centering
  \begin{tikzpicture}[scale=0.735]
    \begin{axis}[
        ybar stacked,
        ymin=0,
        ymax=4,
        xtick=data,
        legend style={cells={anchor=east}, legend pos=north west, legend columns=-1},
        reverse legend=false, % set to false to get correct display, but I'd like to have this true
        xticklabels from table={\optimalvsnot}{Label},
        xticklabel style={text width=2cm,align=center},
        legend plot pos=right,
        ylabel={\Large precision --- prior and posterior},
        xlabel={\Large eigenvector $i$},
      ]
      \addplot [fill=green!80]  table [y=prior,   meta=Label, x expr=\coordindex] {\optimalvsnot};
      \addplot [fill=blue!60]   table [y=optimal, meta=Label, x expr=\coordindex] {\optimalvsnot};     
      \addlegendentry{\Large $\sigma^2\lambda_i^{-1}$}
      \addlegendentry{\Large optimal $\eta_i$s}
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}[scale=0.735]
    \begin{axis}[
        ybar stacked,
        ymin=0,
        ymax=4,
        xtick=data,
        legend style={cells={anchor=east}, legend pos=north west, legend columns=-1},
        reverse legend=false, % set to false to get correct display, but I'd like to have this true
        xticklabels from table={\optimalvsnot}{Label},
        xticklabel style={text width=2cm,align=center},
        legend plot pos=right,
        ylabel={\Large precision --- prior and posterior},
        xlabel={\Large eigenvector $i$},
      ]   
      \addplot [fill=green!80]  table [y=prior,       meta=Label, x expr=\coordindex] {\optimalvsnot};
      \addplot [fill=blue!60]   table [y=sub-optimal, meta=Label, x expr=\coordindex] {\optimalvsnot};
      \addlegendentry{\Large $\sigma^2\lambda_i^{-1}$}
      \addlegendentry{\Large sub-optimal $\eta_i$s}
    \end{axis}
  \end{tikzpicture}
  \caption{A comparison of the eigenvalues of the pushforward
    posterior precision $(\fwd\prcov\fwd^*)^{-1} +
    \sigma^{-2}\obs^*\obs$ for a D-optimal design (left) and a
    sub-optimal design (right). Both designs are allowed $m=3$
    measurements. We assume $\sigma^2=1$ and thus, the blue area has
    accumulated height of $\sigma^{-2}m = 3$ in both panels. The
    D-optimal design (left) increases precision where it is
    lowest. The sub-optimal design (right) does not.}
  \label{fig:optimal vs not}
\end{figure}
